Master Thesis

### 2013-05-15 ig.tex
Wrote about the IG-ACG link

>>>>

Both abstract categorial grammars and interaction grammars, having come
out of the same research team, are both closely related to linear
logic. Not only are these two formalisms based on the same logical
framework, there is a surprisingly direct connection between the
principles guiding the syntactic composition of constants having
intuitionistic implicative linear types in abstract categorial grammars
and of polarized tree descriptions in interaction grammars.

Perrier \cite{perrier1999intuitionistic} proved an interesting result
which bridges these two worlds. His theorem states that every IILL
(intuitionistic implicative linear logic) sequent $F_1, ..., F_n \vdash
G$ is provable if and only if the syntactic description $D((F_1 \limp
... \limp F_n \limp G)^+)$ is valid. The second premise deserves some
elaboration.

What we mean by a \emph{syntactic description} is something similar to a
polarized tree description, albeit more minimalistic. Syntactic
descriptions only talk about immediate dominance and (large) dominance,
there are no precedence constraints on sister nodes and (large)
dominance cannot be restricted by a feature structure. Nodes themselves
are no longer even represented by feature structures, but by simple
atomic categories paired with polarities. Every node in a syntactic
description is thus associated with a single atomic category and a
polarity which is either positive or negative.

A syntactic description that is \emph{valid} is one for which there
exists a model. A model of a syntactic description is defined in the
same manner as for polarized tree descriptions, but instead of unifying
feature structures, we simply demand that all nodes of the description
which map to a node in the model must have the same category and the
number of positively polarized nodes must equal the number of negatively
polarized nodes.

Finally, we explain the term $D((F_1 \limp ... \limp F_n \limp
G)^+)$. The $+$ operation recursively assigns positive and negative
polarities to all formula occurrences within $F_1 \limp ... \limp F_n
\limp G$. These polarities then drive the recursive definition of $D$,
which maps this polarized formula into a syntactic description. Since we
will not cover the theorem deeply in our treatise, we will settle for
knowing that the composition of the positive polarization operation $+$
and the syntactic description producing operation $D$ maps an IILL
formula into a syntactic description such that above-stated theorem
holds. Furthermore, this mapping can be easily constructed since Perrier
gives straightforward ways of computing the results of both operations.

This theorem establishes a striking symmetry between the two
formalisms. Let us consider the case of parsing a sentence using a
lexicalized grammar in both formalism. First, the sentence is tokenized
into wordforms and for each wordform, a lexical item is selected, be it
a typed constant in some abstract signature for ACGs or an EPTD for IGs.

In ACGs, we will try to take the typed constants and to use them to
produce a term having some distinguished type $s$. Thanks to the
Curry-Howard correspondence, constructing a term having a given type is
nothing more than proving that term's type as a formula in the logic of
our type system. What this means is that our parsing problem boils down
to proving the IILL sequent $\tau_1, ..., \tau_n \vdash s$ (we omit the
inclusion of intuitionistic non-linear implications since they are not
relevant for parsing sentences). The sentence thus becomes parseable by
our grammar under the given lexical selection if and only if this
sequent is provable and furthermore, the proof of the sequent is our
desired syntactic structure in a disguise.

Now in IGs, we will conjoin all the EPTDs into a single complex
polarized tree description. Parsing is then just a matter of finding a
model for this description. The sentence is therefore parseable by our
grammar under the given lexical selection if and only if this
description is valid and furthermore, the model which proves the
description valid is our desired syntactic structure.

This symmetry turns Perrier's theorem into a strong claim that lets us
transform the types of a signature into elementary syntactic
descriptions such that some selection of descriptions will generate some
tree if and only if the selection of the corresponging typed constants
generated some properly-typed term. Inverting this transformation would
then let us take the elementary syntactic descriptions of an IG like
Frigram and turn them into types for an abstract signature of our ACG.

However, this approach is not sufficient. The $D \circ +$ transformation
is far from a bijection.

Firstly, it is not injective and several types end up being represented
by the same syntactic description. What this means is that the
transformation ignores the insignificant differences between formulas
such as $a \limp (b \limp c)$ and $b \limp (a \limp c)$. This would not
pose a serious issue though, as when we would try to invert $D \circ +$,
we could simply choose one of the possible formulas which fit the
syntactic description as our canonical representation.

A bigger problem is posed by the fact that $D \circ +$ is not
surjective. Perrier \cite{perrier2001intuitionistic} characterized the
set of syntactic descriptions that can be obtained as images of IILL
sequents using our transformation $D \circ +$, the set of \emph{IILL
  tree descriptions}, which is a proper subset of the set of all
syntactic descriptions. Specifically, in an IILL tree descriptions, all
immediate dominance constraints go from negative nodes to positive
nodes, all (large) dominance constraints lead from positive nodes to
negative nodes and the root has a positive polarity.

While this property may not hold for the PTDs we find in Frigram, it
still gives us a general plan for recovering IILL types. The strategy
suggested by the description-formula connection given by the theorem
yields a mapping $F$ from the under-specified trees to IILL formulas.

$$
F(N) = F(N_1) \limp ... \limp F(N_k) \limp cat(N)
$$

where $N_1, ..., N_k$ are the children of $N$ and $cat(N)$ is the
category of $N$. If $N$ is a leaf, then $F(N)$ is simply $cat(N)$. In
this definition, we treat the tree descriptions as trees with the union
of the immediate dominance and (large) dominance constraints serving as
the edges of the tree.

This strategy corresponds to our intuitions about tree descriptions
should be expressed using IILL types. We will consider two very simple
cases to demonstrate it.

First, picture a simple tree description with a positive root $A$ and
two (necessarily) negative children $B$ and $C$. Our strategy would give
this tree the type $B \limp C \limp A$ (or $C \limp B \limp A$), which
correctly expresses the fact that this tree needs to consume some node
of category $B$ and another one of category $C$ and is able to provide a
node of category $A$. This tree description can also be easily
translated to TAG by turning the negative leaf nodes into substitution
nodes. The conventional translation of TAG into categorial grammar then
confirms the type proposed by our strategy.

The second example will demonstrate the translation of a Frigram
EPTD. Before we do that, however, we will first discuss some of the ways
the actual PTDs of interaction grammars and Frigram differ from the IILL
tree descriptions we just talked about.

The PTDs of IG include precedence constraints. We can see them as
serving two purposes, to both provide an ordering of the constituents
and to further constrain syntactic composition. In our translation to an
ACG, we could imagine handling the first role, that of providing word
order, by defining a lexicon from our abstract signature. Each constant
would combine its arguments in the order in which their corresponding
subtrees appeared in the ordered PTD. However, the second role of
precedence constraints, to establish further restrictions on syntactic
composition, would be lost, as the only restricting factor on syntactic
composition in ACGs is the type system.

Another large difference is the use of multiple independent polarized
features per node. A possible solution here is to salvage the most
salient polarized feature, $cat$ (possibly with its complementary
feature, $funct$), and ignore the others. The results and intuitions we
have about intepreting tree descriptions as types do not offer an easy
way of modeling polarities across different dimensions, features, at the
same time. This would mean that phenomena, such as the paired
grammatical words for negation in French, which were handled using
special polarized features will need new solutions in the framework of
ACGs.

Next up are the virtual features. They serve multiple purposes as
well. Most often they are used to select nodes in the context and to
link them to the nodes contributed by the EPTD.

A common pattern using virtual features is Frigram's handling of
modifiers. The way we would model a modifier in categorial grammars
would be to have a function of type $X \limp X$ where $X$ is the type of
the consituent being modified. In TAG, we would have an adjunction tree
with a root and a foot node of category $X$ and which pins the modifier
on the $X$ phrase. This approach would work in interaction grammars as
well but the formalism offers other solutions too. The designers of
Frigram have opted for an alternate treatment in which modifying a
constituent does not increase the depth of the parse tree. Modifiers are
adjoined as sister nodes of the constituents they modify and it is
through virtual features that IGs can ensure that a modifier is inserted
only in the proper context.

See Figure~\ref{fig:ig-adj} for an example of an adjective modifier in
Frigram. Nodes (1,3) and (1,2) select the noun to be modified and its
projection, respectively. Singling out this pair of nodes lets us hang
the adjective as a child of the projection and state that it must
precede the noun in the word order. We can see that this does not
increase the depth of the noun in the parse tree, since the noun is only
referred to using virtual features in this EPTD, meaning that some other
EPTD must provide the nodes that will saturate these virtual features.

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{ig-adj.pdf}
  \caption{\label{fig:ig-adj} An EPTD for the adjective $grand$ in
    Frigram.}
\end{figure}

Understanding this pattern can give us a useful tool for reading
EPTDs. Consider the EPTD of $la$ we saw in
Figure~\ref{fig:ig-eptds}. The EPTD can now be seen as applying this
pattern twice at the same time to both cliticize the pronoun to the
verbal kernel and to fill the object valency slot of the
verb.

Another use of virtual features in IGs is to restrict the range of
(large) dominance relations. We have seen this before in the EPTD of the
determiner $aucun$ (Figure~\ref{fig:ig-neg}), which had to reach up to
the nearest containing clause and hang a negative $neg$ feature there.
It is also present in the EPTD of the relative pronoun $que$
(Figure~\ref{fig:ig-que}) which will bring us back to the translation
strategy proposed above.

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{ig-que.pdf}
  \caption{\label{fig:ig-que} An EPTD of the relative pronoun $que$.}
\end{figure}

Once again, at the root of the EPTD we see the modifier pattern in
action. If we take the subtree formed by the nodes (2,8) and (2,9) to
represent a modification of some constituent, say a noun, then we get a
type like $n \limp n$. However, the EPTD contains also positive and
negative $cat$ polarities. The node (2,0) has a negative $cat$ polarity
and we can consider it as an argument, meaning that the type we would
like to derive would look like $F((2,0)) \limp (n \limp n)$. Our
strategy $F$ has nothing to say about virtual features or (large)
dominance constraints. If we look only at the positive and negative
$cat$ features, $F((2,0))$ becomes $np \limp s$, which would give us
$(np \limp s) \limp (n \limp n)$ as the final type.

However, this translation is very lossy in terms of its discriminatory
power. Many of the constraints which were forced by the original EPTD
are now ignored. We can imagine boosting our strategy by using the
feature structures as our types instead of atomic categories, which
would not only ensure that e.g. the $sent\_type$ of the embedded clause
must be $decl$, but would also capture the information in the polarized
$funct$ feature so that the relative pronoun $que$ is not permitted to
extract subjects. There are still constraints that we would miss out
though, such as the fact that the subject (2,3) of the clause (2,0) must
precede any embedded clauses (2,6)...(2,5), the fact that any such
embedded clause must serve either the $modal$ or the $obj$ function and
the fact that the embedded clause (2,5) must have a subject (2,4).

Other constraints that we did not mention also do not have a direct
equivalent in ACGs. One example would be a feature of IGs that we have
not even mentioned which allows us the grammar author to provide an
\emph{exhaustive} list of children of a node in a PTD. Another kind of
tricky constraint are the orange rectangles that we can see in
Figure~\ref{fig:ig-que} which enforce that a constituent must be the
left-most or right-most daughter. The problem with these is similar as
with precedence relations. To model their impact on word order would be
easy using an ACG lexicon, but to model their effect on restricting
syntactic composition would be more tricky.

In the end, even though both formalisms (IGs and ACGs) are built on the
framework of linear logic and their fundamental principles of syntactic
composition have been linked together by Perrier's theorem, we conclude
that embedding Frigram into an ACG or constructing some automated
technique for translating the Frigram EPTDs into ACG types and terms
would not be feasible. Perhaps using some of the other connectives of
linear logic would allows us to model the Frigram EPTDs, and the ways
they manipulate multiple resources at the same time, in a more efficient
manner. However, this would lead us well out of the framework of ACGs in
which we wish to conduct our research.

In our grammar, we will still gladly use the lexicon Frilex, which
proves the usefulness of clearly defining the interface between the
lexicon from the grammar in the architecture of Frigram. In our grammar,
Frigram can still serve as a guide on how to interface with Frilex, on
which phenomena need to be covered and in what detail, and what is a
good way to structure the reusable components of a grammar of French.
